---
title: "Longitudinal analysis - RNA expression"
author: "Anna Costa-Garrido"
date: "2024-05-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```



```{r}
# Packages
library(dplyr)
library(biomaRt)
library(tibble)
library(tidyr)
library(ggplot2)

# library(kmlShape)
library(kml)
# library(TMixClust)

library(GOxploreR)
library(rrvgo)
```


# Identify clusters of genes according to the temporal evolution (differentiation process) of their expression

## Exploratory: using hierarchical clustering

- With protein coding genes that are differentially expressed across the B-cell development
N = 12.576 genes
```{r}
load("/home/acost1/BALL_project/results/limma/limma_results_sv.rds")

selected_genes <- unique(c(
results$prepro %>% filter(padj < 0.01) %>% rownames(),
results$proB %>% filter(padj < 0.01) %>% rownames(),
results$preB %>% filter(padj < 0.01) %>% rownames(),
results$transB %>% filter(padj < 0.01) %>% rownames(),
results$nB %>% filter(padj < 0.01) %>% rownames(),
results$gcb %>% filter(padj < 0.01) %>% rownames(),
results$memB %>% filter(padj < 0.01) %>% rownames(),
results$plasma %>% filter(padj < 0.01) %>% rownames() ) ) 

# 17.743
```




```{r}
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
genes <- biomaRt::getBM(filters= "ensembl_gene_id",
                          attributes = c("ensembl_gene_id", "transcript_biotype"), 
                          values = selected_genes, 
                          mart = mart)
  
genes$protein_coding_c <- ifelse(genes$transcript_biotype == "protein_coding", 1, 0)
genes <- genes %>% dplyr::select(ensembl_gene_id, protein_coding_c) %>%
                   group_by(ensembl_gene_id) %>%
                   summarise(protein_coding_c = sum(protein_coding_c)) %>%
                   filter(protein_coding_c == 1)

selected_genes <- selected_genes[selected_genes %in% genes$ensembl_gene_id]
# 12.576 genes
```

```{r}
load("/home/acost1/BALL_project/data/RNA/counts_rlog_normalized.rds")

NormByRlog_selected <- NormByRlog[selected_genes, !sub("\\_.*", "", colnames(NormByRlog)) %in%
                                    c("CMP", "Mon", "nCD8")] 
rm(NormByRlog)
```


```{r}
# Euclidean distance
dist <- dist(NormByRlog_selected, diag=TRUE)

# Hierarchical Clustering with hclust
hc <- hclust(dist)

# Plot the result
plot(hc)
abline(h = 20, col = "brown", lwd = 2)


gene_cluster <- cutree(hc, k = 50) %>% 
  # turn the named vector into a tibble
  enframe() %>% 
  # rename some of the columns
  rename(gene = name, cluster = value)


counts_long <- NormByRlog_selected %>%
               as.data.frame() %>%
               mutate(gene = rownames(NormByRlog_selected) )%>%
               pivot_longer(!gene,names_to = "cell_type", values_to = "count")
counts_long$group_cell_type <- sub("\\_.*", "", counts_long$cell_type)
counts_long <- group_by(counts_long,gene, group_cell_type) %>%
               summarise(avg_counts = mean(count)) %>%
               filter(!group_cell_type %in% c("CMP", "Mon", "nCD8")) 

counts_cluster <- counts_long %>% 
                  inner_join(gene_cluster, by = "gene")

counts_cluster$group_cell_type <- factor(counts_cluster$group_cell_type,
                                         levels = c("HSC", "PreProB", "ProB", "PreB", "immtransB", "nB", "GCB", "memB", "PC"),
                                         labels = c("HSC", "PreProB", "ProB", "PreB", "immtransB", "nB", "GCB", "memB", "PC"))

counts_cluster %>% 
  ggplot(aes(group_cell_type, avg_counts)) +
  geom_line(aes(group = gene)) +
  geom_line(stat = "summary", fun = "median", colour = "brown", size = 1.5, 
            aes(group = 1)) +
  facet_wrap( ~ cluster, ncol = 6)
ggsave("/home/acost1/BALL_project/results/longitudinal/clusters.png", width = 50, height = 100, units = "cm")
```



```{r}
library("pheatmap")
pheatmap(NormByRlog_selected, cutree_rows = 13,
          scale = 'row')
```



## KMLshape

```{r}
traj <- rename(counts_long, id = gene, times = group_cell_type, traj = avg_counts)
traj <- traj[order(traj$id, traj$times), ]
traj$id <- as.integer(traj$id)
traj$times <- as.integer(traj$times)

myClds <- cldsLong(as.data.frame(traj))
plot(myClds)


kml <- kmlShape(myClds, nbClusters = 50, timeScale = 0.1, FrechetSumOrMax = "max", 
         toPlot="both", parAlgo=parKmlShape())
```


## KmL

```{r}
counts_long$group_cell_type <- factor(counts_long$group_cell_type,
                                         levels = c("HSC", "PreProB", "ProB", "PreB", "immtransB", "nB", "GCB", "memB", "PC"),
                                         labels = c("HSC", "PreProB", "ProB", "PreB", "immtransB", "nB", "GCB", "memB", "PC"))


counts_long <- arrange(counts_long, gene, group_cell_type)

# Pivot long to wide 
counts_wide <- pivot_wider(counts_long,
                           id_cols = "gene", 
                           values_from = "avg_counts",
                           names_from = "group_cell_type")
counts_wide <- as.data.frame(counts_wide)
rownames(counts_wide) <- counts_wide$gene
counts_wide <- counts_wide[,-1]


counts_cluster <- clusterLongData(traj=counts_wide,idAll=rownames(counts_wide))

clusters <- kml(counts_cluster,nbClusters=10:20, nbRedrawing=20, toPlot="criterion", 
                parAlgo=parALGO() )
    
### Mean without individual

png("/home/acost1/BALL_project/results/longitudinal/plot_trajectories13.png", 
    height = 22, width = 40, units = "cm", res = 300)
plot(counts_cluster, 13,  parTraj=parTRAJ(type="n"))
dev.off()

png("/home/acost1/BALL_project/results/longitudinal/plot_trajectories20.png", 
    height = 22, width = 40, units = "cm", res = 300)
plot(counts_cluster, 20,  parTraj=parTRAJ(type="n"))
dev.off()



plot(counts_cluster, 13, parMean=parMEAN(type="l"))
### Only one letter on the mean trajectories
plot(counts_cluster, 13, parMean=parMEAN(pchPeriod=Inf))
### Color individual according to its clusters (col="clusters")
plot(counts_cluster, 13, parTraj=parTRAJ(col="clusters"))




plot(counts_cluster, 20, parTraj=parTRAJ(col="clusters"),parMean=parMEAN(type="n"))
### Only few trajectories
plot(counts_cluster, 20, nbSample=1,parTraj=parTRAJ(col='clusters'),parMean=parMEAN(type="n"))


plotTraj(counts_cluster,20)
```


```{r}
genes_cluster13 <- data.frame(gene = rownames(counts_wide), cluster = getClusters(counts_cluster,13) )
genes_cluster20 <- data.frame(gene = rownames(counts_wide), cluster = getClusters(counts_cluster,20) )
                              
counts_long %>% inner_join(genes_cluster13, by = "gene") %>%
  ggplot(aes(group_cell_type, avg_counts)) +
  geom_line(aes(group = gene)) +
  geom_line(stat = "summary", fun = "mean", colour = "brown", size = 1.5, 
            aes(group = 1)) +
  facet_wrap( ~ cluster, ncol = 6) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ggsave("/home/acost1/BALL_project/results/longitudinal/ggplot_trajectories13.png",
       height = 22, width = 50, units = "cm") 

                             
counts_long %>% inner_join(genes_cluster20, by = "gene") %>%
  ggplot(aes(group_cell_type, avg_counts)) +
  geom_line(aes(group = gene)) +
  geom_line(stat = "summary", fun = "mean", colour = "brown", size = 1.5, 
            aes(group = 1)) +
  facet_wrap( ~ cluster, ncol = 6) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ggsave("/home/acost1/BALL_project/results/longitudinal/ggplot_trajectories20.png",
       height = 22, width = 50, units = "cm")

```



Heatmap to show the rna expression of the genes along the stages

```{r}
load("/home/acost1/BALL_project/results/longitudinal/kml_results.rds") 
genes_cluster13 <- data.frame(gene = rownames(counts_wide), cluster = getClusters(counts_cluster,13) )


example_genes <- genes_cluster13 %>% group_by(cluster) %>% slice(1:5)

NormByRlog_selected2 <- NormByRlog_selected[example_genes$gene, ]

counts_long_selected <- NormByRlog_selected2 %>%
               as.data.frame() %>%
               mutate(gene = rownames(NormByRlog_selected2) )%>%
               pivot_longer(!gene,names_to = "cell_type", values_to = "count")
counts_long_selected$group_cell_type <- sub("\\_.*", "", counts_long_selected$cell_type)

counts_long_selected <- merge(counts_long_selected, genes_cluster13, by = "gene", all.x = TRUE)

samples_ordered <- c("HSC_fetal_1", "HSC_fetal_2", "HSC_fetal_3",
                     "PreProB_fetal_1", "PreProB_fetal_2", "PreProB_fetal_3",
                     "ProB_fetal_1", "ProB_fetal_2", "ProB_fetal_3",
                     "PreB_fetal_1", "PreB_fetal_2", "PreB_fetal_3",
                     "immtransB_fetal_1", "immtransB_fetal_2", "immtransB_fetal_3",
                     "nB_WT_1", "nB_WT_2", "nB_WT_3",
                     "GCB_tonsil_1", "GCB_tonsil_3", "GCB_tonsil_4", "GCB_tonsil_5",
                     "memB_WT_1", "memB_WT_2", "memB_WT_3", "memB_WT_4", 
                     "PC_WT_1", "PC_WT_2", "PC_WT_3" )


# Heatmap 
counts_long_selected$cell_type <- factor(counts_long_selected$cell_type,
                                               levels = samples_ordered,
                                               labels = samples_ordered)
counts_long_selected$group_cell_type <- factor(counts_long_selected$group_cell_type,
                                               levels = c("HSC", "PreProB", "ProB", "PreB", "immtransB", "nB", "GCB", "memB", "PC"),
                                               labels = c("HSC", "PreProB", "ProB", "PreB", "immtransB", "nB", "GCB", "memB", "PC"))


counts_long_selected %>% 
  ggplot(aes(cell_type, count, colour = gene)) +
  geom_line(aes(group = gene)) +
  facet_wrap( ~ cluster, ncol = 6) + 
  theme(legend.position = "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))



ggplot(counts_long_selected, aes(x = cell_type, y = gene, fill = count)) +
  geom_tile() +
  facet_grid(cluster~., scales="free", space = "free")  +
   scale_fill_gradient2(mid="#eee8d5", high="#dc322f", low="#268bd2") + 
  theme(strip.text.y = element_text(angle = 0)) +
  xlab("") + ylab("")
```



Enrichment analysis - ORA of each cluster (GO and ReactomePA)

```{r}
load("/home/acost1/BALL_project/results/longitudinal/kml_results.rds")

counts_cluster

genes_cluster13 <- data.frame(gene = rownames(counts_wide), cluster = getClusters(counts_cluster,13) )



enrichment_clusters <- apply(data.frame(names(table(genes_cluster13$cluster))), 1 , function(x){ 
enrichGO(gene = genes_cluster13$gene[genes_cluster13$cluster == x],
                   universe = genes_cluster13$gene,
                   ont ="ALL", 
                   keyType = "ENSEMBL", 
                   readable = T, 
                   pvalueCutoff = 1, 
                   OrgDb = org.Hs.eg.db, 
                   pAdjustMethod = "fdr") %>% as.data.frame() } )
clusterA %>% as.data.frame()
```


```{r}
lapply(enrichment_clusters, function(x){ 
  x <- filter(x, ONTOLOGY == "BP")
  c(x %>% nrow(), filter(x, pvalue < 0.05) %>% nrow(),  filter(x, p.adjust < 0.05) %>% nrow() )
})



# 2. Get the 20 most important BP terms from the each contrast (sorted by the p-value) 
# Only BP
results_BP_terms <- lapply(enrichment_clusters[-5], function(x){ 
  x <- filter(x, ONTOLOGY == "BP" & p.adjust < 0.05)
  aux <- data.frame(id = x$ID, description = x$Description, pvalue = -log10(x$p.adjust))
  aux <- aux[ order(rev(aux$pvalue))  , ]
  levels <- GOTermBPOnLevel(goterm = aux$id)
  aux <- merge(aux, levels, by.x = "id", by.y = "Term", all.x = TRUE)
  aux <- filter(aux, Level >= 4 & Level <= 10)
  aux$id[1:20]
 }
)  
results_BP_terms <- unique(Reduce(c,results_BP_terms) ) # n = 38 (1006)


names(enrichment_clusters) <- paste0("cluster", 1:13)

# 3. Get the enrichment score for these terms in each contrast
results_BP <-  lapply(enrichment_clusters, function(x) {
              x <- filter(x, ONTOLOGY == "BP" & p.adjust < 0.05)
              aux <- data.frame(id = x$ID, description = x$Description,  pvalue = -log10(x$p.adjust))
              aux <- aux[ order(rev(aux$pvalue))  , ]
              aux <- aux[aux$id %in% results_BP_terms, ]
              dplyr::select(aux, id, description, pvalue) } ) %>%
              purrr::reduce(full_join, by = c("id", "description") )

names(results_BP) <- c("id", "description", names(enrichment_clusters))



# 4. Calculate the similarity matrix for these terms using REVIGO and find clusters
simMatrix <- calculateSimMatrix(results_BP_terms,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")
reducedTerms <- reduceSimMatrix(simMatrix, scores = NULL, threshold=0.7,
                                orgdb="org.Hs.eg.db")


results_BP <- merge(results_BP, 
                    dplyr::select(reducedTerms, go, parentTerm),
                    by.x = "id", by.y = "go", all.x = TRUE)

colnames(results_BP) <- c("id", "description", names(enrichment_clusters), "cluster")

results_BP <- results_BP[order(results_BP$cluster, decreasing = TRUE),]


# 5. Plot with heatmap()
results_BP_matrix <- as.matrix(results_BP[,colnames(results_BP) %in% names(enrichment_clusters) ] )
rownames(results_BP_matrix) <- results_BP$description




# Plot with ggplot2()
results_BP_matrix <- reshape2::melt(results_BP_matrix)
results_BP_matrix <- merge(results_BP_matrix, 
                           dplyr::select(results_BP, description, cluster),
                           by.x = "Var1", by.y = "description", all.x = TRUE)


ggplot(results_BP_matrix, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  facet_grid(cluster~., scales="free", space = "free")  +
   scale_fill_gradient2(midpoint = 0, mid="#eee8d5", high="#dc322f", low="#268bd2") + 
  theme(strip.text.y = element_text(angle = 0)) +
  xlab("") + ylab("") + labs(fill="Enrichment score") + ggtitle("Biological Process")
```


```{r}
pathways_clusters <- apply(data.frame(names(table(genes_cluster13$cluster))), 1 , function(x){ 
enrichPathway(gene = getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id",
                                                "entrezgene_id"),values=genes_cluster13$gene[genes_cluster13$cluster == x],mart= mart)$entrezgene_id,
                  pvalueCutoff = 1, readable=TRUE
              ) %>% as.data.frame() } )
clusterA %>% as.data.frame()



lapply(pathways_clusters, function(x) filter(x, p.adjust < 0.01) )
```


## TMixClust

```{r}
plot_time_series_df(counts_wide[1:500,])

cluster_obj = TMixClust(counts_wide[1:100,], nb_clusters = 30)


for (i in 1:10) {
  # extract the time series in the current cluster and plot them
  c_df=counts_wide[which(cluster_obj$em_cluster_assignment==i),]
  plot_time_series_df(c_df, plot_title = paste("cluster",i))
}


plot_silhouette(cluster_obj)



# Other methods
  # - BClustLonG
  # - mfuzz

```




# Clustering with mclust

```{r}
library(mclust)
```


Clustering on the scaled and normalized counts by rlog
```{r}
mod1 <- Mclust(t(scale(t(NormByRlog_selected))), modelNames = "EII")
summary(mod1)
table(mod1$classification)


genes_cluster9 <- data.frame(gene = rownames(NormByRlog_selected), cluster = mod1$classification )
  

counts_long <- t(scale(t(NormByRlog_selected))) %>%
               as.data.frame() %>%
               mutate(gene = rownames(NormByRlog_selected) )%>%
               pivot_longer(!gene,names_to = "cell_type", values_to = "count")
counts_long$group_cell_type <- sub("\\_.*", "", counts_long$cell_type)
counts_long <- group_by(counts_long,gene, group_cell_type) %>%
               filter(!group_cell_type %in% c("CMP", "Mon", "nCD8")) 


counts_long_selected <- counts_long %>% inner_join(genes_cluster9, by = "gene")


counts_long_selected$cell_type <- factor(counts_long_selected$cell_type,
                                         levels = colnames(NormByRlog_selected),
                                         labels = colnames(NormByRlog_selected))

ggplot( counts_long_selected , aes(x = cell_type, y = gene, fill = count)) +
  geom_tile() +
  facet_grid(cluster~., scales="free", space = "free")  +
  scale_fill_gradient2(mid="#eee8d5", high="#dc322f", low="#268bd2") + 
  theme(strip.text.y = element_text(angle = 0),
        axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab("") + ylab("")


counts_long_selected %>%
  ggplot(aes(cell_type, count)) +
  geom_line(aes(group = gene)) +
  geom_line(stat = "summary", fun = "mean", colour = "brown", size = 1.5, 
            aes(group = 1)) +
  facet_wrap( ~ cluster, ncol = 6) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```


Clustering on the normalized counts by Median of ratio's - DESeq2

```{r}
load("/home/acost1/BALL_project/results/limma/limma_results_sv.rds")


# Selected genes from DE analysis of limma with SV
selected_genes <- unique(c(
results$prepro %>% filter(padj < 0.01 & abs(log2FoldChange) >= 3) %>% rownames(),
results$proB %>% filter(padj < 0.01 & abs(log2FoldChange) >= 3) %>% rownames(),
results$preB %>% filter(padj < 0.01 & abs(log2FoldChange) >= 3) %>% rownames(),
results$transB %>% filter(padj < 0.01 & abs(log2FoldChange) >= 3) %>% rownames(),
results$nB %>% filter(padj < 0.01 & abs(log2FoldChange) >= 3) %>% rownames(),
results$gcb %>% filter(padj < 0.01 & abs(log2FoldChange) >= 3) %>% rownames(),
results$memB %>% filter(padj < 0.01 & abs(log2FoldChange) >= 3) %>% rownames(),
results$plasma %>% filter(padj < 0.01 & abs(log2FoldChange) >= 3) %>% rownames() ) )  
# N = 9,229


# Select protein-coding genes
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
genes <- biomaRt::getBM(filters= "ensembl_gene_id",
                          attributes = c("ensembl_gene_id", "transcript_biotype"), 
                          values = selected_genes, 
                          mart = mart)
  
genes$protein_coding_c <- ifelse(genes$transcript_biotype == "protein_coding", 1, 0)
genes <- genes %>% dplyr::select(ensembl_gene_id, protein_coding_c) %>%
                   group_by(ensembl_gene_id) %>%
                   summarise(protein_coding_c = sum(protein_coding_c)) %>%
                   filter(protein_coding_c == 1)

selected_genes <- selected_genes[selected_genes %in% genes$ensembl_gene_id]
# 5,316 genes



load("/home/acost1/BALL_project/data/RNA/counts_deseq2_normalized.rds")

normalized_selected <- normalized_counts[selected_genes,] 
rm(normalized_counts)


# Clustering with mclust
mod1 <- Mclust(t(scale(t(normalized_selected))), modelNames = "EII")
summary(mod1)
table(mod1$classification)


genes_cluster4 <- data.frame(gene = rownames(normalized_selected), cluster = mod1$classification )
  

counts_long <- t(scale(t(normalized_selected))) %>%
               as.data.frame() %>%
               mutate(gene = rownames(normalized_selected) )%>%
               pivot_longer(!gene,names_to = "cell_type", values_to = "count")
counts_long$group_cell_type <- sub("\\_.*", "", counts_long$cell_type)
counts_long <- group_by(counts_long,gene, group_cell_type) %>%
               filter(!group_cell_type %in% c("CMP", "Mon", "nCD8")) 


counts_long_selected <- counts_long %>% inner_join(genes_cluster4, by = "gene")


counts_long_selected$cell_type <- factor(counts_long_selected$cell_type,
                                         levels = colnames(normalized_selected),
                                         labels = colnames(normalized_selected))

ggplot( counts_long_selected , aes(x = cell_type, y = gene, fill = count)) +
  geom_tile() +
  facet_grid(cluster~., scales="free", space = "free")  +
  scale_fill_gradient2(mid="#eee8d5", high="#dc322f", low="#268bd2") + 
  theme(strip.text.y = element_text(angle = 0),
        axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab("") + ylab("")


counts_long_selected %>%
  ggplot(aes(cell_type, count)) +
  geom_line(aes(group = gene)) +
  geom_line(stat = "summary", fun = "mean", colour = "brown", size = 1.5, 
            aes(group = 1)) +
  facet_wrap( ~ cluster, ncol = 6) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))



save(mod1, file = "/home/acost1/BALL_project/results/predict_expression/group_mclust.rds")
```

Enrichment analysis

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
```


```{r}
enrichment_clusters <- apply(data.frame(names(table(genes_cluster4$cluster))), 1 , function(x){ 
                  enrichGO(gene = genes_cluster4$gene[genes_cluster4$cluster == x],
                   universe = genes_cluster4$gene,
                   ont ="ALL", 
                   keyType = "ENSEMBL", 
                   readable = T, 
                   pvalueCutoff = 1, 
                   OrgDb = org.Hs.eg.db, 
                   pAdjustMethod = "fdr") %>% as.data.frame() } )
```



```{r}
lapply(enrichment_clusters, function(x){ 
  x <- filter(x, ONTOLOGY == "BP")
  c(x %>% nrow(), filter(x, pvalue < 0.05) %>% nrow(),  filter(x, p.adjust < 0.05) %>% nrow() )
})



# 2. Get the 20 most important BP terms from the each contrast (sorted by the p-value) 
# Only BP
results_BP_terms <- lapply(enrichment_clusters[-c(5,6,9)], function(x){ 
  x <- filter(x, ONTOLOGY == "BP" & p.adjust < 0.05)
  aux <- data.frame(id = x$ID, description = x$Description, pvalue = -log10(x$p.adjust))
  aux <- aux[ order(rev(aux$pvalue))  , ]
  levels <- GOTermBPOnLevel(goterm = aux$id)
  aux <- merge(aux, levels, by.x = "id", by.y = "Term", all.x = TRUE)
  aux <- filter(aux, Level >= 4 & Level <= 10)
  aux$id[1:20]
 }
)  
results_BP_terms <- unique(Reduce(c,results_BP_terms) ) # n = 38 (1006)


names(enrichment_clusters) <- paste0("cluster", 1:9)

# 3. Get the enrichment score for these terms in each contrast
results_BP <-  lapply(enrichment_clusters[-c(5,6,9)], function(x) {
              x <- filter(x, ONTOLOGY == "BP" & p.adjust < 0.05)
              aux <- data.frame(id = x$ID, description = x$Description,  pvalue = -log10(x$p.adjust))
              aux <- aux[ order(rev(aux$pvalue))  , ]
              aux <- aux[aux$id %in% results_BP_terms, ]
              dplyr::select(aux, id, description, pvalue) } ) %>%
              purrr::reduce(full_join, by = c("id", "description") )

names(results_BP) <- c("id", "description", names(enrichment_clusters)[-c(5,6,9)])



# 4. Calculate the similarity matrix for these terms using REVIGO and find clusters
simMatrix <- calculateSimMatrix(results_BP_terms,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")
reducedTerms <- reduceSimMatrix(simMatrix, scores = NULL, threshold=0.7,
                                orgdb="org.Hs.eg.db")


results_BP <- merge(results_BP, 
                    dplyr::select(reducedTerms, go, parentTerm),
                    by.x = "id", by.y = "go", all.x = TRUE)

colnames(results_BP) <- c("id", "description", names(enrichment_clusters)[-c(5,6,9)], "cluster")

results_BP <- results_BP[order(results_BP$cluster, decreasing = TRUE),]


# 5. Plot with heatmap()
results_BP_matrix <- as.matrix(results_BP[,colnames(results_BP) %in% names(enrichment_clusters)[-c(5,6,9)] ] )
rownames(results_BP_matrix) <- results_BP$description




# Plot with ggplot2()
results_BP_matrix <- reshape2::melt(results_BP_matrix)
results_BP_matrix <- merge(results_BP_matrix, 
                           dplyr::select(results_BP, description, cluster),
                           by.x = "Var1", by.y = "description", all.x = TRUE)


ggplot(results_BP_matrix, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  facet_grid(cluster~., scales="free", space = "free")  +
   scale_fill_gradient2(midpoint = 0, mid="#eee8d5", high="#dc322f", low="#268bd2") + 
  theme(strip.text.y = element_text(angle = 0)) +
  xlab("") + ylab("") + labs(fill="Enrichment score") + ggtitle("Biological Process")
ggsave("/home/acost1/BALL_project/results/gsea/gsea_bp_clusters.png",
       height = 35, width = 65, units = "cm")
```



```{r}
pathways_clusters <- apply(data.frame(names(table(genes_cluster4$cluster))), 1 , function(x){ 
                            enrichPathway(gene = getBM(filters= "ensembl_gene_id", 
                                                   attributes= c("ensembl_gene_id", "entrezgene_id"),
                                                   values = genes_cluster4$gene[genes_cluster4$cluster == x], 
                                                   mart= mart)$entrezgene_id,
                                      pvalueCutoff = 1, readable=TRUE) %>% as.data.frame() } )

lapply(pathways_clusters, function(x) filter(x, p.adjust < 0.01) )
```


# Clustering of the genes depending on the results obtained with Limma - SV

```{r}
load("/home/acost1/BALL_project/results/limma/limma_results_sv.rds")

counter <- 0 
contrast <- names(results[-1])
results_limma_sv <- lapply(results[-1], function(x) {
                    counter <<- counter + 1
                    x <- x[order(rownames(x)), ]
                    x$cell_type <- "constant"
                    x$cell_type[x$padj < 0.01 & x$log2FoldChange > 3] <- "up"
                    x$cell_type[x$padj < 0.01 & x$log2FoldChange < -3] <- "down"
                    x <- dplyr::select(x, cell_type) %>%
                           mutate(gene_id = rownames(x)) 
                    colnames(x)[1] <- contrast[counter]
                    x
                    } 
                    ) %>% purrr::reduce( full_join, by = "gene_id")
results_limma_sv[is.na(results_limma_sv)] <- "constant"
```



```{r}
results_limma_sv <- data.frame(
  gene_id = results_limma_sv$gene_id,
  group = apply(results_limma_sv, 1, function(x) paste0(x[-2], collapse = "-") ) )


table(results_limma_sv$group) %>% length() # n = 426 groups 

sort(table(results_limma_sv$group), decreasing = TRUE)[1:30] %>% as.data.frame()


# Most common
most_common <- table(results_limma_sv$group) %>% sort(decreasing = TRUE) %>% head(n = 30) %>% names()
results_limma_sv_common <- results_limma_sv[results_limma_sv$group %in% most_common,]
# n genes = 20.204



# Get the expression
selected_genes <- results_limma_sv_common$gene_id

mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
genes <- biomaRt::getBM(filters= "ensembl_gene_id",
                          attributes = c("ensembl_gene_id", "transcript_biotype"), 
                          values = selected_genes, 
                          mart = mart)
  
genes$protein_coding_c <- ifelse(genes$transcript_biotype == "protein_coding", 1, 0)
genes <- genes %>% dplyr::select(ensembl_gene_id, protein_coding_c) %>%
                   group_by(ensembl_gene_id) %>%
                   summarise(protein_coding_c = sum(protein_coding_c)) %>%
                   filter(protein_coding_c == 1)

selected_genes <- selected_genes[selected_genes %in% genes$ensembl_gene_id]
# n protein coding genes = 13.130


load("/home/acost1/BALL_project/data/RNA/counts_rlog_normalized.rds")

NormByRlog_selected <- NormByRlog[selected_genes, !sub("\\_.*", "", colnames(NormByRlog)) %in%
                                    c("CMP", "Mon", "nCD8")] 



counts_long <- NormByRlog_selected %>%
               as.data.frame() %>%
               mutate(gene = rownames(NormByRlog_selected) )%>%
               pivot_longer(!gene,names_to = "cell_type", values_to = "count")
counts_long$group_cell_type <- sub("\\_.*", "", counts_long$cell_type)
counts_long <- group_by(counts_long,gene, group_cell_type) %>%
               summarise(avg_counts = mean(count)) %>%
               filter(!group_cell_type %in% c("CMP", "Mon", "nCD8")) 


counts_long$group_cell_type <- factor(counts_long$group_cell_type,
                                         levels = c("HSC", "PreProB", "ProB", "PreB", "immtransB", "nB", "GCB", "memB", "PC"),
                                         labels = c("HSC", "PreProB", "ProB", "PreB", "immtransB", "nB", "GCB", "memB", "PC"))


counts_long <- arrange(counts_long, gene, group_cell_type)

counts_long %>% merge(results_limma_sv_common, by.x = "gene", by.y = "gene_id", all.y = TRUE) %>%
  filter(!is.na(group_cell_type)) %>%
  ggplot(aes(group_cell_type, avg_counts)) +
  geom_line(aes(group = gene)) +
  geom_line(stat = "summary", fun = "mean", colour = "brown", size = 1.5, 
            aes(group = 1)) +
  facet_wrap( ~ group, ncol = 6) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ggsave("/home/acost1/BALL_project/results/longitudinal/ggplot_trajectories.png",
       height = 30, width = 70, units = "cm")
```


```{r}
# Enrichment analysis
x <- enrichPathway(gene=de, pvalueCutoff = 0.05, readable=TRUE)


genes_cluster <- counts_long %>% merge(results_limma_sv_common, by.x = "gene", by.y = "gene_id", all.y = TRUE)
genes_cluster <- dplyr::select(genes_cluster, gene, group) %>%
                        rename(cluster = group) 
  


pathways_clusters2 <- apply(data.frame(names(table(genes_cluster$cluster))), 1 , function(x){ 
enrichPathway(gene = getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id",
                                                "entrezgene_id"),
                           values=genes_cluster$gene[genes_cluster$cluster == x],mart= mart)$entrezgene_id,
                  pvalueCutoff = 1, readable=TRUE
              )  } )




pathways_clusters_sig <- lapply(pathways_clusters, function(x) filter(x, p.adjust < 0.01) )

names(pathways_clusters_sig) <- data.frame(names(table(genes_cluster$cluster)))[,1]

for(i in 1:30){enrichplot::cnetplot(pathways_clusters2[[i]], showCategory=20) }


enrichplot::cnetplot(pathways_clusters2[[1]])
enrichplot::treeplot(pairwise_termsim(pathways_clusters2[[1]]), hclust_method = "average")     
enrichplot::treeplot(pairwise_termsim(pathways_clusters2[[3]]),  hclust_method = "average"  )
enrichplot::treeplot(pairwise_termsim(pathways_clusters2[[4]]),  hclust_method = "average"  )
enrichplot::treeplot(pairwise_termsim(pathways_clusters2[[5]]),  hclust_method = "average"  )
enrichplot::treeplot(pairwise_termsim(pathways_clusters2[[6]]),  hclust_method = "average"  )
enrichplot::treeplot(pairwise_termsim(pathways_clusters2[[7]]),  hclust_method = "average"  )
enrichplot::treeplot(pairwise_termsim(pathways_clusters2[[8]]),  hclust_method = "average"  )
```