---
title: "Network analysis"
author: "Anna CG"
date: "2024-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

```{r}
# Packages
library(igraph)
library(dplyr)
library(Matrix)
library(gplots)

library(pheatmap)
library(RColorBrewer)
```


```{r}
# Load data
load("/home/acost1/BALL_project/results/gnn/seq_bed_file_notsplit.rds")
load("/home/acost1/BALL_project/results/gnn/edge_file_notsplit.rds")
```


```{r}
# Create igraph network
network_celltype <- function(cell_type_select){
  edge_file <- filter(edge_file,  cell_type == cell_type_select)
  id_bed <- filter(id_bed, cell_type == cell_type_select)
  
  links <- data.frame(
        source = edge_file$ID_1,
        target = edge_file$ID_2
        )

  nodes <- data.frame(
              name = id_bed$id,
              carac = id_bed$type ) %>%
             filter(!duplicated(.))

  # Turn it into igraph object
  network <- graph_from_data_frame(d=links, vertices=nodes, directed=F)
}
```

```{r}
all_cell_types <- names(table(id_bed$cell_type))
all_networks <- apply(data.frame(all_cell_types), 1, function(x) network_celltype(x) )
```

```{r}
length(unique(id_bed$id))
length(unique(V(all_networks[[2]])$name ))
```



1. Node and edge conservation

```{r}
# Calculate conserved nodes across stages
conserved_OE <- Reduce(intersect, lapply(all_networks, function(g) V(g)$name[V(g)$carac == "OE"]))
all_OE <- unique(sapply(all_networks, function(g) V(g)$name[V(g)$carac == "OE"]) %>% unlist)

length(conserved_OE)
length(all_OE)

conserved_B <- Reduce(intersect, lapply(all_networks, function(g) V(g)$name[V(g)$carac == "B"]))
all_B <- unique(sapply(all_networks, function(g) V(g)$name[V(g)$carac == "B"]) %>% unlist)

length(conserved_B)
length(all_B)

# Calculate conserved edges across stages
edge_list <- lapply(all_networks, get.edgelist) %>%
              lapply(., function(x) apply(x, 1, paste, collapse = "-") )
all_edges <- unique(edge_list %>% unlist)
conserved_edges <- Reduce(intersect, edge_list)
print("Conserved edges across all stages:")
print(length(conserved_edges))
print(length(all_edges))
```


Pairwaise matrix node conservation
```{r}
# Initialize a matrix to store node overlap counts / jaccard
B_overlap_matrix <- matrix(0, nrow = 9, ncol = 9)
OE_overlap_matrix <- matrix(0, nrow = 9, ncol = 9)

jaccard_similarity_B <- matrix(0, nrow = 9, ncol = 9)
jaccard_similarity_OE <- matrix(0, nrow = 9, ncol = 9)

# Calculate node overlap between each pair of stages
for (i in 1:(9 - 1)) {
  for (j in (i + 1):9) {
    # Find common nodes between stages i and j
    nodes_B_i <- V(all_networks[[i]])$name[V(all_networks[[i]])$carac == "B"]
    nodes_B_j <- V(all_networks[[j]])$name[V(all_networks[[i]])$carac == "B"]
    
    common_B <- intersect(nodes_B_i, nodes_B_j)
    
    nodes_OE_i <- V(all_networks[[i]])$name[V(all_networks[[i]])$carac == "OE"]
    nodes_OE_j <- V(all_networks[[j]])$name[V(all_networks[[i]])$carac == "OE"]
    
    common_OE <- intersect(nodes_OE_i, nodes_OE_j)
    
    # Store the count of common nodes
    B_overlap_matrix[i, j] <- length(common_B)
    B_overlap_matrix[j, i] <- B_overlap_matrix[i, j] # Symmetric matrix
    
    OE_overlap_matrix[i, j] <- length(common_OE)
    OE_overlap_matrix[j, i] <- OE_overlap_matrix[i, j] # Symmetric matrix
    
    
    # Jaccard
    intersection <- length(common_B)
    union <- length(union(nodes_B_i, nodes_B_j))
    jaccard_similarity_B[i, j] <- intersection / union
    jaccard_similarity_B[j, i] <- jaccard_similarity_B[i, j] # Symmetric matrix
    
    intersection <- length(common_OE)
    union <- length(union(nodes_OE_i, nodes_OE_j))
    jaccard_similarity_OE[i, j] <- intersection / union
    jaccard_similarity_OE[j, i] <- jaccard_similarity_OE[i, j] # Symmetric matrix
    
  }
}

# Add stage labels
colnames(B_overlap_matrix) <- all_cell_types
rownames(B_overlap_matrix) <- all_cell_types

colnames(OE_overlap_matrix) <- all_cell_types
rownames(OE_overlap_matrix) <- all_cell_types

palette <- colorRampPalette(brewer.pal(n = 7, name = "Greens"))(100)

id_bed_B <- filter(id_bed, type == "B")
diag(B_overlap_matrix) <- tapply(id_bed_B$id, id_bed_B$cell_type, function(x) length(unique(x)))

id_bed_OE <- filter(id_bed, type == "OE")
diag(OE_overlap_matrix) <- tapply(id_bed_OE$id, id_bed_OE$cell_type, function(x) length(unique(x)))

# Plot heatmap

breaks <- seq(0, 1, length.out = 100)


png("/home/acost1/BALL_project/results/networks/bait_conservation_number.png", width = 50, height = 25, res = 300, units = "cm")
pheatmap(B_overlap_matrix,
        cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.0f",
         color = palette,
         border_color = NA,
         fontsize = 15,
         main = "Bait Conservation Heatmap",
         fontsize_row = 15,
         fontsize_col = 15)
dev.off()

png("/home/acost1/BALL_project/results/networks/OE_conservation_number.png", width = 50, height = 25, res = 300, units = "cm")
pheatmap(OE_overlap_matrix,
        cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.0f",
         color = palette,
         border_color = NA,
         fontsize = 15,
         main = "OE Conservation Heatmap",
         fontsize_row = 15,
         fontsize_col = 15)
dev.off()

colnames(jaccard_similarity_B) <- all_cell_types
rownames(jaccard_similarity_B) <- all_cell_types

colnames(jaccard_similarity_OE) <- all_cell_types
rownames(jaccard_similarity_OE) <- all_cell_types

diag(jaccard_similarity_B) <- 1
diag(jaccard_similarity_OE) <- 1


breaks <- seq(0, 1, length.out = 100)

png("/home/acost1/BALL_project/results/networks/bait_conservation_jaccard.png", width = 50, height = 25, res = 300, units = "cm")
pheatmap(jaccard_similarity_B, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.2f",
         breaks = breaks,
         number_color  = "black",
         color = palette,
         border_color = NA,
         fontsize = 15,
         main = "Bait Conservation Heatmap (Jaccard Index)",
         fontsize_row = 15,
         fontsize_col = 15)
dev.off()


png("/home/acost1/BALL_project/results/networks/oe_conservation_jaccard.png", width = 50, height = 25, res = 300, units = "cm")
pheatmap(jaccard_similarity_OE, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.2f",
         breaks = breaks,
         color = palette,
         border_color = NA,
         fontsize = 15,
         number_color  = "black",
         main = "OE Conservation Heatmap (Jaccard Index)",
         fontsize_row = 15,
         fontsize_col = 15)
dev.off()
```

Pairwaise matrix interactions conservation
```{r}
# Initialize a matrix to store node overlap counts
interaction_overlap_matrix <- matrix(0, nrow = 9, ncol = 9)

# Calculate interaction overlap between each pair of stages
for (i in 1:(9 - 1)) {
  for (j in (i + 1):9) {
    # Find common interactions between stages i and j
    common_interactions <- intersect(edge_list[[i]], edge_list[[j]])
    # Store the count of common interactions
    interaction_overlap_matrix[i, j] <- length(common_interactions)
    interaction_overlap_matrix[j, i] <- interaction_overlap_matrix[i, j] # Symmetric matrix
  }
}

# Optionally, normalize overlap by total nodes to get a proportion
# node_overlap_matrix <- node_overlap_matrix / num_nodes

# Add stage labels
colnames(interaction_overlap_matrix) <- all_cell_types
rownames(interaction_overlap_matrix) <- all_cell_types

diag(interaction_overlap_matrix) <- sapply(edge_list, length)

# Plot heatmap
png("/home/acost1/BALL_project/results/networks/interaction_conservation_number.png", width = 50, height = 25, res = 300, units = "cm")
pheatmap(interaction_overlap_matrix,
        cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.0f",
         color = palette,
         border_color = NA,
         fontsize = 15,
         main = "Interaction Conservation Heatmap",
         fontsize_row = 15,
         fontsize_col = 15)
dev.off()

# Interaction Conservation: Compute pairwise Jaccard similarity index
jaccard_similarity <- matrix(0, nrow = 9, ncol = 9)
rownames(jaccard_similarity) <- all_cell_types
colnames(jaccard_similarity) <- all_cell_types

for (i in 1:9) {
  interactions_i <- edge_list[[i]]
  for (j in 1:9) {
    interactions_j <- edge_list[[j]]
    intersection <- length(intersect(interactions_i, interactions_j))
    union <- length(union(interactions_i, interactions_j))
    jaccard_similarity[i, j] <- intersection / union
  }
}

# 2. Plot Heatmap of Interaction Conservation

# Plot using pheatmap
png("/home/acost1/BALL_project/results/networks/interaction_conservation_jaccard.png", width = 50, height = 25, res = 300, units = "cm")
pheatmap(jaccard_similarity, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.2f",
         breaks = breaks,
         color = palette,
         border_color = NA,
         fontsize = 15,
         main = "Interaction Conservation Heatmap (Jaccard Index)",
         fontsize_row = 15,
         fontsize_col = 15)
dev.off()
```


2. Community Detection and Comparison


```{r}
network_celltype <- function(cell_type_select){
  edge_file <- filter(edge_file,  cell_type == cell_type_select)
  id_bed <- select(id_bed, id, type) %>% filter(!duplicated(.))
  
  links <- data.frame(
        source = edge_file$ID_1,
        target = edge_file$ID_2
        )

  nodes <- data.frame(
              name = id_bed$id,
              carac = id_bed$type ) %>%
             filter(!duplicated(.))

  # Turn it into igraph object
  network <- graph_from_data_frame(d=links, vertices=nodes, directed=F)
}


all_networks <- apply(data.frame(all_cell_types), 1, function(x) network_celltype(x) )
```


```{r}
# Apply Louvain community detection for each network
communities <- lapply(all_networks, function(g) cluster_louvain(g))

# Calculate NMI between each pair of stages to assess community similarity
nmi_matrix <- matrix(0, 9, 9)
ari_matrix <-  matrix(0, 9, 9)
for (i in 1:(9 - 1)) {
  for (j in (i + 1):9) {
    nmi_matrix[i, j] <- compare(communities[[i]], communities[[j]], method = "nmi")
    nmi_matrix[j, i] <- nmi_matrix[i, j]
    
    ari_matrix[i, j] <- compare(communities[[i]], communities[[j]], method = "adjusted.rand")
    ari_matrix[j, i] <- ari_matrix[i, j]
  }
}
print("NMI Matrix for community similarity between stages:")
print(nmi_matrix)
print(ari_matrix)


# # Function to get common subgraph based on shared nodes
# get_common_subgraph <- function(g1, g2) {
#   common_nodes <- intersect(V(g1)$name, V(g2)$name)
#   sub_g1 <- induced_subgraph(g1, vids = V(g1)$name %in% common_nodes)
#   sub_g2 <- induced_subgraph(g2, vids = V(g2)$name %in% common_nodes)
#   return(list(sub_g1, sub_g2))
# }
# 
# # Applying community detection and comparing communities for each pair
# num_stages <- length(all_networks) # Assuming 'networks' is a list of the stage graphs
# nmi_matrix <- matrix(0, num_stages, num_stages)
# rownames(nmi_matrix) <- all_cell_types
# colnames(nmi_matrix) <- all_cell_types
# 
# for (i in 1:(num_stages - 1)) {
#   for (j in (i + 1):num_stages) {
#     # Get common subgraphs for stages i and j
#     subgraphs <- get_common_subgraph(all_networks[[i]], all_networks[[j]])
#     sub_g1 <- subgraphs[[1]]
#     sub_g2 <- subgraphs[[2]]
#     
#     # Run community detection on the subgraphs
#     comm1 <- cluster_louvain(sub_g1)
#     comm2 <- cluster_louvain(sub_g2)
#     
#     # Compare communities using NMI
#     nmi_matrix[i, j] <- compare(comm1, comm2, method = "nmi")
#     nmi_matrix[j, i] <- nmi_matrix[i, j]  # Mirror the value for symmetry
#   }
# }
# 
# print("NMI Matrix for community similarity between stages:")
# print(nmi_matrix)

rownames(nmi_matrix) <- all_cell_types
colnames(nmi_matrix) <- all_cell_types

diag(nmi_matrix) <- 1

png("/home/acost1/BALL_project/results/networks/community_conservation_nmi.png", width = 50, height = 25, res = 300, units = "cm")
pheatmap(nmi_matrix, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.2f",
         number_color = "white",
         breaks = breaks,
         color = palette,
         border_color = NA,
         fontsize = 15,
         main = "Normalized Mutual Information (NMI) Heatmap",
         fontsize_row = 15,
         fontsize_col = 15)
dev.off()

rownames(ari_matrix) <- all_cell_types
colnames(ari_matrix) <- all_cell_types

diag(ari_matrix) <- 1

png("/home/acost1/BALL_project/results/networks/community_conservation_ari.png", width = 50, height = 25, res = 300, units = "cm")
pheatmap(ari_matrix, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.2f",
         number_color = "black",
         breaks = breaks,
         color = palette,
         border_color = NA,
         fontsize = 15,
         main = "Adjusted Rand Index (ARI) Heatmap",
         fontsize_row = 15,
         fontsize_col = 15)
dev.off()
```


Identify stable and dynamic clusters across the stages

```{r}
cluster_assignments <- lapply(communities, function(x) membership(x) ) %>% bind_cols() 
colnames(cluster_assignments) <- all_cell_types
rownames(cluster_assignments) <- names(membership(communities[[1]]))
apply(cluster_assignments, 2, max) # Number of clusters/communities for each stages

apply(cluster_assignments, 2, function(x){ table_stage <- table(x) 
                                           table_stage[table_stage != 1] %>% length()
})
```



```{r}
table_HSC <- table(cluster_assignments$HSC)
table_PreProB <- table(cluster_assignments$`PrePro-B`)

clusters_HSC <- names(table_HSC[table_HSC > 20])
clusters_PreProB <- names(table_PreProB[table_PreProB > 20])

similar_clusters <- matrix(NA, nrow = length(clusters_HSC), ncol = length(clusters_PreProB))

jaccard_similarity <- function(cluster1, cluster2) {
  intersection <- length(intersect(cluster1, cluster2))
  union <- length(unique(c(cluster1, cluster2)))
  return(intersection / union)
}


for(i in 1: length(clusters_HSC) ){
  for(j in 1:length(clusters_PreProB) ){
  names_genes <- rownames(cluster_assignments)
  
  stages_HSC <- cluster_assignments[,"HSC"]
  stages_PrePro <-  cluster_assignments[,"PrePro-B"]
  
  nodes_in_cluster1 <- names_genes[stages_HSC == clusters_HSC[i] ]
  nodes_in_cluster2 <- names_genes[stages_PrePro == clusters_PreProB[j]  ]

  similar_clusters[i, j] <- ifelse(jaccard_similarity(nodes_in_cluster1, nodes_in_cluster2) > 0.7, 1, 0)
  }
}

which(similar_clusters == 1, arr.ind = TRUE)


stages_HSC[stages_HSC == clusters_HSC[i] ]

clusters_HSC[4] # 24
clusters_PreProB[6] # 60


jaccard_similarity(names(cluster_assignments$HSC[cluster_assignments$HSC == "24"]),
                   names(cluster_assignments$`PrePro-B`[cluster_assignments$`PrePro-B` == "60"])  )



stages_HSC <- cluster_assignments[,"HSC"]
stages_PrePro <-  cluster_assignments[,"PrePro-B"]
  
nodes_in_cluster1 <- which(stages_HSC == 4)
nodes_in_cluster2 <- which(stages_PrePro == 6)

jaccard_similarity(nodes_in_cluster1, nodes_in_cluster2)

length(clusters_HSC)

length(clusters_PreProB)

dim(similar_clusters)
```



```{r}


# Initialize a list to store cumulative Jaccard scores for each stage's clusters
n_stages <- 9
cumulative_jaccard_scores <- vector("list", n_stages)

# Loop over each stage
for (stage in 1:n_stages) {
  stage_clusters <- unique(cluster_assignments[[stage]])
  stage_jaccard_scores <- matrix(0, nrow = length(stage_clusters), ncol = n_stages)
  
  # Loop through each cluster in the current stage
  for (cluster_id in stage_clusters) {
    nodes_in_current_cluster <- which(cluster_assignments[[stage]] == cluster_id)
    
    # Calculate Jaccard similarity across all other stages
    jaccard_scores <- c()
    for (compare_stage in 1:n_stages) {
      if (compare_stage != stage) {
        for (compare_cluster_id in unique(cluster_assignments[[compare_stage]])) {
          nodes_in_compare_cluster <- which(cluster_assignments[[compare_stage]] == compare_cluster_id)
          jaccard_score <- jaccard_similarity(nodes_in_current_cluster, nodes_in_compare_cluster)
          jaccard_scores <- c(jaccard_scores, jaccard_score)
        }
      }
    }
    
    # Store the average Jaccard similarity for this cluster across stages
    stage_jaccard_scores[cluster_id, stage] <- mean(jaccard_scores)
  }
  
  cumulative_jaccard_scores[[stage]] <- stage_jaccard_scores
}

# Display cumulative Jaccard scores for each cluster across stages
colnames(cumulative_jaccard_scores) <- paste("Stage", 1:n_stages, sep = "_")
rownames(cumulative_jaccard_scores) <- paste("Cluster", 1:n_clusters, sep = "_")
print(cumulative_jaccard_scores)
```



Node participation ??

```{r}
tracked_nodes <- c("200320") # Specify nodes to track



# Data structure to store community membership information
node_communities <- list() # Will hold community memberships for each tracked node across stages

# Run community detection and record node community membership
for (stage in 1:num_stages) {
  g <- all_networks[[stage]]
  community <- cluster_louvain(g)
  
  # Record community for each tracked node
  node_communities[[stage]] <- sapply(tracked_nodes, function(node) {
    if (node %in% V(g)$name) {
      V(g)$name[community$membership == membership(community)[V(g)$name == node] ]
    } else {
      NA  # Use NA for nodes that don’t exist in this stage
    }
  })
}

# Convert list to a data frame for easier analysis
node_communities_df <- do.call(rbind, node_communities)
colnames(node_communities_df) <- tracked_nodes
rownames(node_communities_df) <- all_cell_types

print("Community memberships of tracked nodes across stages:")
print(node_communities_df)

# 1. Calculate community stability: Check if nodes stay in the same community across stages
node_stability <- sapply(tracked_nodes, function(node) {
  communities <- node_communities_df[, node]
  communities <- communities[!is.na(communities)]  # Remove NAs (if node is missing in a stage)
  length(unique(communities)) == 1  # TRUE if node stayed in the same community across stages
})

print("Node community stability across stages (TRUE = stable, FALSE = switches):")
print(node_stability)

# 2. Count community switches for each node across stages
node_switches <- sapply(tracked_nodes, function(node) {
  communities <- node_communities_df[, node]
  communities <- communities[!is.na(communities)]
  sum(diff(communities) != 0)  # Counts the number of community switches
})

print("Number of community switches for each node:")
print(node_switches)
```



3. Network Topology Metrics


```{r}
# Calculate degree distribution, density, and centrality for each network
network_metrics <- lapply(all_networks, function(g) {
  degree_distribution <- degree(g)
  edge_density <- edge_density(g)
  centralities <- betweenness(g)
  list(degree_distribution = degree_distribution, edge_density = edge_density, centralities = centralities)
})

# Compare degree distribution across stages
degree_distributions <- sapply(network_metrics, function(m) mean(m$degree_distribution))
print("Average degree distribution across stages:")
print(degree_distributions)

# Compare network densities across stages = number of edges divided by the possible edges
densities <- sapply(network_metrics, function(m) m$edge_density)
print("Edge densities across stages:")
print(densities)

# Compare average betweenness centrality across stages
avg_centralities <- sapply(network_metrics, function(m) mean(m$centralities))
print("Average betweenness centrality across stages:")
print(avg_centralities)
```

```{r}

# Degree distribution = number of edges per node
par(mfrow = c(3, 3))
for (i in 1:num_stages) {
  hist(network_metrics[[i]]$degree_distribution, main = paste("Degree Distribution - Stage", all_cell_types[i]),
       xlab = "Degree", ylab = "Frequency", col = "lightblue")
}


barplot(sapply(network_metrics, function(x) x$edge_density), main = "Edge density")



par(mfrow = c(3, 3))
for (i in 1:num_stages) {
  hist(network_metrics[[i]]$centralities, main = paste("Betweenness centrality - Stage", all_cell_types[i]),
       xlab = "Degree", ylab = "Frequency", col = "lightblue")
}
```


I want to identify nodes with high betweenness and closeness centrality across the differentiation

By analyzing conserved or shifting centrality values, I can spot nodes that act as consistent hubs or connectors, as well as those that may take on different roles in each stage.

```{r}
betweenness_centrality <- list()
closeness_centrality <- list()

# Calculate betweenness and closeness centrality for each node in each stage
for (stage in 1:num_stages) {
  g <- all_networks[[stage]]
  
  # Compute betweenness and closeness, setting names to match node IDs
  betweenness <- betweenness(g, normalized = TRUE)
  closeness <- closeness(g, normalized = TRUE)
  
  # Store with consistent naming
  names(betweenness) <- V(g)$name
  names(closeness) <- V(g)$name
  
  # Save results in the list, ensuring all vectors have the same length
  betweenness_centrality[[stage]] <- betweenness
  closeness_centrality[[stage]] <- closeness
}

all_node_ids <- unique(unlist(lapply(all_networks, function(g) V(g)$name)))


# Convert centrality lists to data frames for easier analysis
betweenness_df <- do.call("cbind", lapply(betweenness_centrality, function(x) {
  x[all_node_ids]  # Match to all possible node IDs
}))
closeness_df <- do.call("cbind", lapply(closeness_centrality, function(x) {
  x[all_node_ids]  # Match to all possible node IDs
}))


high_betweenness <- rowSums(betweenness_df > 0.05)  # Counts of stages with high betweenness
high_closeness <- rowSums(closeness_df > 0.2)  # Counts of stages with high closeness


consistent_high_betweenness <- names(high_betweenness[high_betweenness >= num_stages / 2])
consistent_high_betweenness <- consistent_high_betweenness[!is.na(consistent_high_betweenness)]
consistent_high_closeness <- names(high_closeness[high_closeness >= num_stages / 2])
consistent_high_closeness <- consistent_high_closeness[!is.na(consistent_high_closeness)]


betweenness_changes <- apply(betweenness_matrix, 1, function(x) sum(diff(x) != 0))
closeness_changes <- apply(closeness_matrix, 1, function(x) sum(diff(x) != 0))

# Nodes with most centrality shifts
changing_betweenness_nodes <- names(sort(betweenness_changes, decreasing = TRUE)[1:10])
changing_closeness_nodes <- names(sort(closeness_changes, decreasing = TRUE)[1:10])

print("Nodes with frequent betweenness centrality changes:")
print(changing_betweenness_nodes)

print("Nodes with frequent closeness centrality changes:")
print(changing_closeness_nodes)
```

